### 2021-10-14 documents/sections/extensions.tex
various revisions

\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
From a comparative view, this list is not quite complete, as there is also a reflex of the copula \rc{a(p)}, which is analayzed as a postposition by \textcite[199--201]{alves2017arara}; it is however inflected with \setone prefixes and also shows first person \obj{w-} \exref{araap}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a(p)} \qu{to be} in \ikpeng, only \rc{eti} forms are attested inflected for first person \parencite[401]{gildea2018reconstructing}.

@@

The question of how this happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{In fact, the issue of \emph{when} this happened is also open. It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation. Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.}
As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the four verbs in \exref{ptirw} as being reconstructible to \PTar with \rc{w-}; we offer our reconstructed \PTir forms.
In addition, \trio \obj{ei} \qu{to be}, the other copular root, also shows first person \obj{w-} \parencite[294]{triomeira1999}, allowing us to reconstruct a first-person \rc{w-eʔi-} for \PTir.

\pex<ptirw> \PTir \gl{1}\gl{s_a_} \rc{w-} \parencites[112--115]{meira1998proto}[85]{gildea1994akuriyo}\\
\begin{tabular}[t]{@{}llll@{}}
& \PTir & \trio & \akuriyo\\
\qu{go} & \rc{wɨ-tə(mɨ)-} & \obj{wɨ-tə(mɨ)-} & \obj{ə-təmɨ-}/\obj{wɨ-təemɨ-}  \\
\qu{say} & \rc{wɨ-ka-} & \obj{wɨ-ka-} & \obj{wɨ-ka-} \\
\qu{come} & \rc{w-əʔepɨ-} & \obj{w-əepɨ-} & \obj{Ø-eepɨ-} \\
\qu{be} & \rc{w-ae-} & \obj{w-ae-} & Ø\obj{-aʔe-} \\
\end{tabular}
%\a<aku12> \akuriyo \obj{kəʔ-eepɨ} \qu{we came} \parencite[114]{meira1998proto}
%\a<aku13> \akuriyo \obj{(w)i-toka} \qu{I hit him/her} \parencite[86]{gildea1994akuriyo}
%\a<tricome> \trio \obj{n-epɨ} \qu{s/he came} \parencite[114]{meira1998proto}
\xe

For \akuriyo \qu{to go}, \textcite{gildea1994akuriyo} registered a different prefix \obj{wɨ-}, rather than \posscite{meira1998proto} \obj{ə-}.\footnote{The \akuriyo recorded by Gildea potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998}.}
The two forms can be reconciled by a specific idiosyncratic combination of sound changes:
We suggest that \obj{ə} is the outcome of the lowering of the \rc{ɨ} in the prefix \rc{wɨ-}; the same has happened to the vowel in the \gl{1+2} prefix \rc{kɨt-} \exref{exp.aku-159}.
Also, \obj{w-} appears to have been subject to ongoing erosion in \akuriyo, also evidenced by its absence in \qu{to come} and \qu{to be}, but its presence in \qu{to say} \exref{ptirw}.
This erosion is also found in the etymologically related \gl{1}>\gl{3} prefix \exref{exp.aku-164}, as well as in other Cariban languages, like \ikpeng \pcref{sec:pekodian} or \hixka \pcref{sec:waiwaian}.

\pex<exp>
\a<aku-159> \akuriyo \parencite[][114]{meira1998proto}\\
\begingl
\gla kəʔ-eepɨ//
\glb \gl{1+2}-come//
\glft \qu{We came.}//
\endgl
\a<aku-164> \akuriyo \parencite[][86]{gildea1994akuriyo}\\
\begingl
\gla (w)i-toka//
\glb \gl{1}>\gl{3}-hit//
\glft \qu{I hit him/her.}//
\endgl
\xe
%
For the form of the verb, \textcite[114--115]{meira1998proto} reconstructs \PTar \rc{əepɨ} for the first person, and \rc{eepɨ} for the others, based on the idiosyncratic form in \trio and the vowel length in \akuriyo.
\akuriyo (and \carijo) would then have levelled the original distribution of \rc{əepɨ} and \rc{eepɨ}, similar to what we have suggested for Pekodian in \cref{sec:pekodian}.
We agree with this scenario, with the exception that \trio \obj{əepɨ} is clearly cognate with \rc{ət-epɨ} \pcref{sec:come}, meaning that the \PTir form would have been \rc{əʔepɨ}.

@@

\textcite{gildea1994akuriyo} also recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-}.
They are listed in \exref{akumov.aku}.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Thus, it is uncertain whether these verbs can be reconstructed as taking \rc{w-} in \PTir, although its presence in \akuriyo does suggest so.

@@

Finally, there are two irregularly inflected \gl{s_a_} verbs in \trio, \obj{ɨhtə} \qu{to go down} and \obj{weka}/\obj{oeka} \qu{to defecate} \exref{downshit}.
They have \gl{1}\gl{s_a_} markers \obj{p-} and \obj{k-}, which are otherwise entirely unattested in \trio.\footnote{Although both elements also occur in other irregular forms of these verbs \parencite[180, 325, 331]{triomeira1999}.}
At least for \qu{to go down}, the \akuriyo cognate suggests that the irregular first person form \rc{p-ɨhtə-} can be reconstructed to \PTir.
Whatever their origins, they were not affected by the extension of \rc{t-}.

In summary, in \PTir{} -- at the latest -- the \PC \gl{1}>\gl{3} marker \rc{t-} largely replaced \gl{1}\gl{s_a_} \rc{w-}.
Four verbs are solidly reconstructible as preserving \rc{w-} in \PTir, and \rc{ɨhtə} \qu{to go down} with irregular \rc{p-} was also not affected.
There are four movement verbs in \akuriyo which appear to not have been affected by the extension of \rc{t-}, as well as another irregular verb \obj{weka} \qu{to defecate} in \trio, with first person \obj{k-oeka-}.

\subsection{\akuriyo \obj{k-}}
\label{sec:akuriyo}
After the split-up of \PTir, when \rc{t-} had largely replaced \rc{w-}, \akuriyo innovated yet another \gl{1}\gl{s_a_} marker: \obj{k-}.
It seems to have replaced \rc{t-} only in specific places, with the two markers showing a clear phonologically conditioned distribution in the \akuriyo data available to us \parencite{gildea1994akuriyo}, with all relevant verbs shown in \Cref{tab:aku1sa}.
\textcite[107]{meira1998proto} largely confirms the distribution shown here, but mentions \dbqu{several cases of first person \obj{t-} in \akuriyo{}} (on \obj{ə}-initial verbs), albeit without any examples.
He also suggests that \obj{k-} could be more recent, with which we agree -- since phonologically distributed \rc{t-}/\rc{t͡ʃ-} is reconstructible to \PTir, a scenario whereby \obj{k-} replaces \rc{t-}, but not \rc{t͡ʃ-}, is the most straightforward.
The few \obj{t-} mentioned by \textcite{meira1998proto} were then either reintroduced under \trio influence, or are the last remnants of the replacement of \rc{t-}.

@@

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \exref{carpar}.

\ex<carpar> \carijo \parencites[106]{meira1998proto}[173]{robayo2000avance}\\
\begin{tabular}[t]{@{}lll@{}}
& \obj{tuda} \qu{to arrive} & \obj{eharaga} \qu{to dance}\\
\gl{1} & \obj{ji-tuda-} & \obj{j-eharaga-}\\
\gl{2} & \obj{mɨ-tuda-} & \obj{m-egaraga-}\\
\gl{1+2} & \obj{kɨsi-tuda-} & \obj{kɨs-eharaga-}\\
\gl{3} & \obj{ni-tuda-} & \obj{n-eharaga-}\\
\end{tabular}
\xe
%
Thus, also detransitivized verbs like \obj{ehɨnəhɨ} \qu{to fight}, which are \gl{s_a_} verbs in other languages, take \obj{j-} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

\ex<car-30>\carijo \parencite[][79]{koch1908hiana}\\
\begingl
\glpreamble hẹnẹ(x)tónoko-mā́ɽẹ y-e̱-hẹnẹ́(x)yai̯//
\gla hɨnəhtono-ko=marə j-e-hɨnəh-jai//
\glb enemy-\gl{pl}=with \gl{1}-\gl{detrz}-kill-\gl{npst}.\gl{cert}//
\glft \qu{I fight with the enemies.}//
\endgl
\xe
%

@@

As noted in \cref{sec:taranoan}, this extension also erased any traces of a putative \PTar \gl{1}\gl{s_a_} marker \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.

>>>>

\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
From a comparative view, this list is not quite complete, as there is also a reflex of the copula \rc{a[p]}, which is analayzed as a postposition by \textcite[199--201]{alves2017arara}; it is however inflected with \setone prefixes and also shows first person \obj{w-} \exref{araap}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
While there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested inflected for first person \parencite[401]{gildea2018reconstructing}.

@@

The question of how this happened in detail \parencite[107--112]{meira1998proto} still has no answer, although it seems necessary to assume a scenario whereby both \rc{t-} and \rc{w-} for a time occurred on both transitive and intransitive verbs \parencite[112]{meira1998proto}.\footnote{
In fact, even the issue of \emph{when} this happened is open.
It could have happened at the \PTar stage, but the subsequent introduction of \obj{j-} in \carijo \pcref{sec:carijo} would have erased any traces of such an innovation.
Accordingly, \textcite{meira1998proto} hesitates to assign this extension to a specific proto-language.
We take a conservative stance and reconstruct it to \PTir only, but acknowledge the possibility of it happening already in \PTar.}
As for verbs unaffected by the spread of \rc{t-}, \textcite{meira1998proto} lists the first four verbs in \exref{ptirw} as being reconstructible to \PTar with \rc{w-}, for which we offer our reconstructed \PTir forms.
To this list, we add the other copular root \rc{eʔi} \parentext{\PTar \rc{et͡ʃi} \parencite[165]{meira1998proto}}, which has first person \obj{w-} at least in \trio.

\pex<ptirw> \PTir \gl{1}\gl{s_a_} \rc{w-} \parencites[112--115]{meira1998proto}[294]{triomeira1999}\\ %[85]{gildea1994akuriyo}
\begin{tabular}[t]{@{}llll@{}}
& \PTir & \trio & \akuriyo\\
\qu{go} & \rc{wɨ-tə(mɨ)-} & \obj{wɨ-tə(mɨ)-} & \obj{ə-təmɨ-} \\% (\obj{wɨ-təemɨ-})\\
\qu{say} & \rc{wɨ-ka-} & \obj{wɨ-ka-} & \obj{wɨ-ka-} \\
\qu{come} & \rc{w-əʔepɨ-} & \obj{w-əepɨ-} & \obj{Ø-eepɨ-} \\
\qu{be} & \rc{w-ae-} & \obj{w-ae-} & Ø\obj{-aʔe-} \\
& \rc{w-eʔi-} & \obj{w-ei-} & ?-\obj{eʔi-}\\
\end{tabular}
%\a<aku12> \akuriyo \obj{kəʔ-eepɨ} \qu{we came} \parencite[114]{meira1998proto}
%\a<aku13> \akuriyo \obj{(w)i-toka} \qu{I hit him/her} \parencite[86]{gildea1994akuriyo}
%\a<tricome> \trio \obj{n-epɨ} \qu{s/he came} \parencite[114]{meira1998proto}
\xe

We agree with \posscite[113]{meira1998proto} identification of the idiosyncratic \akuriyo first person prefix \obj{ə-} on \qu{to go} as a reflex of \rc{wɨ-}.
Both components of the irregular change \rc{wɨ-} > \obj{ə-} -- loss of \rc{w} and lowering of \rc{ɨ} to \obj{ə} -- are found in other person prefixes \exref{exp}.

\pex<exp>
\a<aku-164> \akuriyo \parencite[][86]{gildea1994akuriyo}\\
\begingl
\gla (w)i-toka//
\glb \gl{1}>\gl{3}-hit//
\glft \qu{I hit him/her.}//
\endgl
\a<aku-159> \akuriyo \parencite[][114]{meira1998proto}\\
\begingl
\gla kəʔ-eepɨ//
\glb \gl{1+2}-come//
\glft \qu{We came.}//
\endgl
\xe
%
%For \akuriyo \qu{to go}, \textcite{gildea1994akuriyo} registered a different prefix \obj{wɨ-}, rather than \posscite{meira1998proto} \obj{ə-}.\footnote{The \akuriyo recorded by Gildea potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998}.}
%The two forms can be reconciled by a specific idiosyncratic combination of sound changes:
%We suggest that \obj{ə} is the outcome of the lowering of the \rc{ɨ} in the prefix \rc{wɨ-}; the same has happened to the vowel in the \gl{1+2} prefix \rc{kɨt-} \exref{exp.aku-159}.
%Also, \obj{w-} appears to have been subject to ongoing erosion in \akuriyo, also evidenced by its absence in \qu{to come} and \qu{to be}, but its presence in \qu{to say} \exref{ptirw}.
%This erosion is also found in the etymologically related \gl{1}>\gl{3} prefix \exref{exp.aku-164}, as well as in other Cariban languages, like \ikpeng \pcref{sec:pekodian} or \hixka \pcref{sec:waiwaian}.
%
%
%
For \qu{to come}, \textcite[114--115]{meira1998proto} reconstructs \PTar \rc{əepɨ} for the first person, and \rc{eepɨ} for the others, based on the idiosyncratic form in \trio and the vowel length in \akuriyo.
\akuriyo (and \carijo) then levelled the original distribution of \rc{əepɨ} and \rc{eepɨ}, similar to what we have suggested for Pekodian \pcref{sec:pekodian}.
We agree with this scenario, with the exception that \trio \obj{əepɨ} is clearly cognate with \rc{ət-epɨ} \pcref{sec:come}, meaning that the \PTir form must have been \rc{əʔepɨ}.

@@

Further, here are two irregularly inflected \gl{s_a_} verbs in \trio, \obj{ɨhtə} \qu{to go down} and \obj{weka}/\obj{oeka} \qu{to defecate} \exref{downshit}.
They have \gl{1}\gl{s_a_} markers \obj{p-} and \obj{k-}, which are otherwise entirely unattested in \trio.\footnote{Although both elements also occur in other irregular forms of these verbs \parencite[180, 325, 331]{triomeira1999}.}
At least for \qu{to go down}, the \akuriyo cognate suggests that the irregular first person form \rc{p-ɨhtə-} can be reconstructed to \PTir.
Whatever their origins, they were not affected by the extension of \rc{t-}.

Finally, \textcite{gildea1994akuriyo} also recorded four more \akuriyo verbs seemingly not affected by innovative \rc{t-} \exref{akumov.aku}, all \obj{e}-initial movement verbs.
We have only found a \trio cognate for \obj{erama} \qu{to return}, which behaves like a regular \gl{s_a_} verb in taking \obj{s-} \exref{akumov.tri}.
Further, these verbs are not mentioned by \textcite{meira1998proto}, who was also working with \posscite{gildea1994akuriyo} data.
Given that this data potentially has strong \trio and/or \wayana influence \parencite[253]{gildea1998} and the lack of support by Meira's data, we cannot reconstruct these verbs as not being affected by the extension of \PTir \rc{t-}.

@@

%In summary, in \PTir{} -- at the latest -- the \PC \gl{1}>\gl{3} marker \rc{t-} largely replaced \gl{1}\gl{s_a_} \rc{w-}.
%Five verbs are solidly reconstructible as preserving \rc{w-} in \PTir, and \rc{ɨhtə} \qu{to go down} with irregular \rc{p-} was also not affected.
%There is another irregular verb \obj{weka} \qu{to defecate} in \trio, with first person \obj{k-oeka-}.

\subsection{\akuriyo \obj{k-}}
\label{sec:akuriyo}
After the split-up of \PTir, when \rc{t-} had largely replaced \rc{w-}, \akuriyo innovated yet another \gl{1}\gl{s_a_} marker: \obj{k-}.
It seems to have replaced \rc{t-} only in specific environments, with the two markers showing a clear phonologically conditioned distribution in the \akuriyo data available to us \parencite{gildea1994akuriyo}, with all relevant verbs shown in \Cref{tab:aku1sa}.
\textcite[107]{meira1998proto} largely confirms the distribution shown here, but mentions \dbqu{several cases of first person \obj{t-} in \akuriyo{}} (on \obj{ə}-initial verbs), albeit without any examples.
He also suggests that \obj{k-} might be more recent, with which we agree -- since the distribution \envr{\rc{t-}}{\obj{ə}} / \envr{\rc{t͡ʃ-}}{e} is reconstructible to \PTir, the most likely scenario is \obj{k-} replacing \rc{t-} but not \rc{t͡ʃ-}.
The few \obj{t-} mentioned by \textcite{meira1998proto} were then either reintroduced under \trio influence, or are the last remnants of the replacement of \rc{t-}.

@@

\subsection{\carijo \obj{j-}}
\label{sec:carijo}
As discussed by \textcite[105--107]{meira1998proto}, \carijo has extended the \gl{1}\gl{s_p_} marker \obj{j(i)-} to \gl{s_a_} verbs.\footnote{Since almost all \gl{s_a_} verbs are V-initial, only the \envr{}{V} allomorph \obj{j-} is attested.}
This, in combination with the extension of \gl{2}\gl{s_a_} \obj{m-} and \gl{1+2}\gl{s_a_} \obj{kɨt-}/\obj{kɨs-} to \gl{s_p_} verbs, results in a single unified \gl{s} category \exref{carpar}.

\ex<carpar> \carijo \parencites[106]{meira1998proto}[173]{robayo2000avance}\\
\begin{tabular}[t]{@{}lll@{}}
& \obj{tuda} \qu{to arrive} & \obj{eharaga} \qu{to dance}\\
\gl{1} & \obj{ji-tuda-} & \obj{j-eharaga-}\\
\gl{2} & \obj{mɨ-tuda-} & \obj{m-eharaga-}\\
\gl{1+2} & \obj{kɨsi-tuda-} & \obj{kɨs-eharaga-}\\
\gl{3} & \obj{ni-tuda-} & \obj{n-eharaga-}\\
\end{tabular}
\xe
%
Although the split-\gl{s} system has been lost entirely, former \gl{s_a_} verbs can be identified by taking a detransitivizer, like \obj{ehɨnəhɨ} \qu{to fight}, derived from \obj{hɨnəhɨ} \qu{to kill} \parencite[179]{robayo2000avance} \exref{car-30}.
%The same marker is also found in \gl{3}>\gl{1} scenarios \exref{car1p.car-1} and possessed nouns \exref{car1p.car-31}.

\ex<car-30>\carijo \parencite[][79]{koch1908hiana}\\
\begingl
\glpreamble hẹnẹ(x)tónoko-mā́ɽẹ y-e̱-hẹnẹ́(x)yai̯//
\gla hɨnəhtono-ko=marə j-e-hɨnəhɨ-jai//
\glb enemy-\gl{pl}=with \gl{1}-\gl{detrz}-kill-\gl{npst}.\gl{cert}//
\glft \qu{I fight with the enemies.}//
\endgl
\xe
%

@@

As noted in \cref{sec:taranoan}, this extension also erased any traces of a potential \PTar \gl{1}\gl{s_a_} marker \rc{t-}.
However, it did not fully eclipse the old \gl{1}\gl{s_a_} marker \rc{w-}, which is attested as being preserved in the verbs \obj{tə} \qu{to go} \exref{car1w.car-24} and \obj{a} \qu{to be} \exref{car1w.car-25}.
